Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-12 paper/type-inference.tex
Type Inference Edits.

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

@@

Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
every definition in the class table $CT$, while accumulating
constraints. The constraints are finally solved by $\solve$
(\S\ref{sec:fb-constraintsolving}), which returns substitution
functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
predicate variables, respectively, introduced during templatization
and elaboration stages. The substitutions are applied to the class
table (and to the artifacts that make up the class table, recursively)
to compute a class table that maps classes to their fully
region-annotated definitions in $\FB$.

@@

\subsection{Constraints}
\label{sec:fb-constraintsem}

@@

\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
template from \S\ref{sec:fb-templatization}. Following constraints are
generated during its elaboration (Constrains are identified with
$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
\rhoenv_0$, where $\rhoenv_0 = \{\rhoalloc_0,\rho_{0-4}\}$, have been
elided): 
\begin{smathpar}
\begin{array}{l}
  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \spc
  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
\end{array}
\end{smathpar}

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts an argument $\C{p}$ of type
$\C{Pair}\inang{\C{A},\C{A}}$, and two objects ($\C{x}$ and $\C{y}$)
of type $A$. It then assigns one object to $\C{p.fst}$ and other to
$\C{p.snd}$, and calls itself recursively without ever terminating.
Which one is assigned to $\C{fst}$, and which to $\C{snd}$, is
alternated between recursive calls. Elaboration phase elaborates
$\C{alt}$ to the following region-annotated definition\footnote{The
original definition of \C{alt} can be obtained by erasing all the
region annotations from the elaborated version}\footnote{The
elaboration actually uses new region variables as parameters to the
recursive call, and then generates constraints that unify them with
the ones shown. We coalesced both steps to simplify presentation.}:
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
                          A<$\rho_9$> x, A<$\rho_{10}$> y) {
  p.fst = x; p.snd = y; 
  this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-10}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{6} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{7} \isvalid{\varphi_2}{\rho_7 \outlives \rho_6 \conj \rho_8
  \outlives \rho_6} \\
\csid{8} \isvalid{\varphi_2}{\rho_9 = \rho_7} \qquad
\csid{9} \isvalid{\varphi_2}{\rho_{10} = \rho_8} \\
\csid{10} \isvalid{\varphi_2}{[\rho_9/\rho_{10}][\rho_{10}/\rho_9]\varphi_2}
\end{array}
\end{smathpar}
% The first constraint is generated by $\typeOk$ on the type of
% $\C{p}$.  Second and third are generated by the the assignment
% expressions, and the last constraint is generated from the recursive
% call.

@@

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints, which are
subsequently solved enmasse. The motivation behind the whole-program
approach to constraint generation is twofold: it simplifies
elaboration functions and makes presentation easier, and second, it
naturally generalizes to mutual recursion. Nonetheless, we do not
intend our type inference to be a whole-program analysis for (a). it
preempts opportunities for separate compilation and dynamic linking,
and (b). it is expensive and an overkill in most practical cases. We
therefore reclaim the compositionality of type inference by solving
the constraints in a compositional fashion. In more practical terms
this means that our constraint solving algorithm visits and solves
every constraint (or, every set of mutually dependent constraints)
only once. It composes computed solutions to solve other constraints
that depend on the solved constraints. Importantly, the failure to
solve a dependent constraint does not result in backtracking. We now
describe our compositional constraint solving algorithm in detail. To
simplify the presentation of our algorithm, we assume that are no
mutually recursive definitions in the source program. Recursive
definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occuring on the left side of the turnstile is said to occur
negatively, or with \emph{negative polarity}. In contrast, a predicate
variable occuring on the right side is said to occur positively, or
with \emph{positive polarity}. A validity constraint
\emph{constrains} the set of predicate variables that occur
negatively in the constraint, while it \emph{uses} the set of
predicate variables that occur positively. A constraint is said to be
\emph{recursive} if it constrains and uses a predicate variable.
% A pair of validity
% constraints are said to be \emph{mutually dependent} if there exists a
% pair ($\varphi_1$, $\varphi_2$) of predicate variables that occur
% with opposite polarities in both the constraints. In such case, the
% pair ($\varphi_1$,$\varphi_2$) of predicate variables are also said to
% be mutually dependent. We call the transitive closure of mutual
% dependency relation as \emph{transitive dependency} relation. Like
% mutual dependency, transitive dependency is also extended to predicate
% variables.

Given a set of validity constraints, we first build a dependency graph
($G_c$) with constraints as nodes, and dependencies between them
captured as edges. There exists an edge from a constraint $c_2$ to a
constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
any of the following conditions hold:
\begin{itemize}
\item $c_1$ constrains a predicate variable that $c_2$ uses. 
\item $c_1$ constrains a (non-strict) subset of predicate variables
that $c_2$ constraint.
\end{itemize}
The first condition intuitively corresponds to a case, where
expression or type, whose region elaboration is constrained by $c_2$
refers to a method or a class, whose unknown precondition is
constrained by $c_1$. The common predicate variable ($\varphi$)
represents the unknown precondition in this case. The dependency from
$c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
before $c_2$ is solved, thus enforcing the rule that the precondition
of a method must not depend on its calling context. 
% Note that the first condition results in self-loops over recursive
% constraints in the dependency graph.  Since mutually recursive
% definitions refer each other, the first condition also results in
% bidirectional dependencies between mutually recursive constraints
% (and self-loops over recursive constraints). 
The second condition captures two kinds of dependencies. First, the
dependency from the constraints over a method precondition to the
constraints over the precondition of the class containing the method.
This captures our preference that the constraints over a class's
region parameters should not depend on the idiosyncracies of its
methods. Any additional constraints required by any of its methods
must be captured in the precondition of the method itself
(well-formedness rules allow this possibility). The second condition
adds bidirectional dependencies between validity constraints that
constraint the same set of predicate variables.

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. Consequently, when
the turn of a constraint set ($C$) arrives during the constraint solving
process, it satisfies certain properties:
\begin{itemize}
\item There exists only one predicate variable ($\varphi$) that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

Solving the set ($C$) of constraints entails finding an assignment for
$C$'s subject ($\varphi$), and also all the region variables
($\rhobar$) that occur free in $C$, such that the solution satisfies
well-formedness constraints on $\varphi$ and $\rhobar$. To simplify
presentation, we think of $C$ as being parameterized on
$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
\rbrack$. We now formalize the constraint satisfaction problem, and
its solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
satisfaction problem is a tuple $\CSP$, where $C\lbrack
\varphi,\rhobar \rbrack$ is a set of validity constraints, where
$i$'th validity constraint assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
unification domain ($\rhoenv$) of the CSP.  The solution to the
constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
constraint formula such that:
\begin{itemize}
\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
$\tywf{\rhoenv_{\varphi}}{\phisol}$).
\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
is valid.
% \begin{center}
% \(
%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
%           {\phicstr^{i} \conj  F_i(\phisol))}
% \)
% \end{center}
\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
\end{itemize}
\end{definition}

\subsection{Solving the CSP}

The first step of solving the CSP $\CSP$ is to cast it as an
equivalent problem $\csp$ involving a single validity constraint. The
single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
\begin{center}
\(
  \isvalid{ \Phicx \conj \varphi}
        {\Phics \conj  \bigwedge_i F_i(\varphi)}
\)
\end{center}
Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
\phicstr^{i}$. To see why both CSPs are equivalent, consider two
distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
generated before $c_j$ by the constraint generation algorithm. Observe
that when our constraint generation algorithm generates a constraint,
all relationships between concrete region identifiers referred by the
consequent of constraint are already present in its antecedent (this
includes the unification domains of region variables referred by the
consequent).  Since no new relationships between existing regions are
added when a new constraints is generated, $\phictxt^j$ either
describes the same relationships that are already described by
$\phictxt^i$, or describes relationships among region identifiers that
are new, and not relevant to $c_i$. Therefore, strengthening the
context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
nor strengthens $c_i$ (resp.  $c_j$).

\subsubsection{Non-Recursive Constraints}
\label{sec:csolve}

We first describe how we solve a non-recursive CSP $\csp$ where
$c\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
\conj \varphi}{\Phics}$.

Our approach is based on the observation that $\FB$ does not admit
null values or uninitialized variables, thus forcing every region
variable to be unified with some concrete region. The constraint
formula $\Phics$ captures all such unification constraints on
$\rhobar$. Since the set of all concrete region identifiers is the
unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
in which case the well-formedness constraint on $\rho_i$ is not
satisfied, and constraint solving must fail.  Therefore, there exists
a unique assignment $\substFn$ such that $c\lbrack
\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
$\phi_{sol}$.

>>>>

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. FGJ type $\RgnZ\inang{T}$ is special; it is
always templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
region-annotated definition of \C{Pair} class from
\S\ref{sec:fb-syntax} a region type template for a method with FGJ
type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
accept one region parameter - the allocation region of their
objects.}\footnote{We abuse arrow notation to also represent types of
methods, but unlike function types, there is no allocation region
annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
region variables, and $\varphi_0$ is a fresh predicate variable
denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
$\rho$ is fresh. For a class, a template is computed in two steps. In
the first step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

@@

Function $\elabClassTable$ (not shown) elaborates
every definition in the class table $CT$, while accumulating
constraints. The constraints are finally solved for free region and
predicate variables introduced during templatization and elaboration
stages. The solutions are applied to the class table (and to the
artifacts that make up the class table, recursively) to compute a
class table that maps classes to their fully region-annotated
definitions in $\FB$.

@@

\subsection{Constraints}
\label{sec:constraints}

@@

\paragraph{Constraints Example 1} Consider the $\C{Pair}$ class
template from \S\ref{sec:fb-templatization}. Following constraints are
generated during its elaboration (Constrains are identified with
$\mathbf{c_i}$'s. Some trivial constraints, such as $\rho_4 \in
\rhoenv_0$ and $\rho_5 \in \rhoenv_1$, where $\rhoenv_0 =
\{\rhoalloc_0,\rho_{0-4}\}$ and $\rhoenv_1 = \rhoenv_0 \cup
\{\rho_5\}$, have been elided): 
\begin{smathpar}
\begin{array}{l}
  \csid{1} \tywf{\rhoenv_0}{\varphi_0} \qquad
  \csid{2} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0} \\
  \csid{3} \isvalid{\varphi_0}{\rho_2 = \rho_0} \spc
  \csid{4} \isvalid{\varphi_0}{\rho_3 = \rho_1} \\
  \csid{5} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\spc
  \csid{6} \tywf{\rhoenv_1}{\varphi_1} \qquad
\end{array}
\end{smathpar}

\paragraph{Constraints Example 2} Let us add to the \C{Pair} class a
contrived method $\C{alt}$ that accepts a \C{Region} object \C{r}, a
\C{Pair<A,A>} object \C{q}, and an \C{A} object \C{y}. It assigns
\C{y} to \C{fst} and \C{snd} fields of \C{q}, and calls itself
recursively with the same region, a new \C{Pair} object allocated in a
local region, and an \C{A} object referred by the \C{snd} field of the
pair inside the region. \C{alt} never terminates.  Elaboration phase
elaborates the method to the following region-annotated
definition\footnote{In reality, elaboration uses new region variables
as parameters to the constructor and method calls, and then generates
constraints that unify them with actuals. In our examples, to avoid
clutter due to trivial constraints,we coalesced both steps and show
the actuals instead.}(The original definition of \C{alt} can be
obtained by erasing all the region annotations from the elaborated
version):
% \begin{codejava}
% unit alt<$\rhoalloc_2$,$\rho_{6-10}$ | $\varphi_2$>(Pair<A,A><$\rho_6$,$\rho_7$,$\rho_8$> p,
%                           A<$\rho_9$> x, A<$\rho_{10}$> y) {
%   p.fst = x; p.snd = y; 
%   this.alt<$\rhoalloc_2$,$\rho_7$,$\rho_8$,$\rho_{10}$,$\rho_9$>(A,y,x);
% }
% \end{codejava}
\begin{codejava}
unit alt<$\rhoalloc_2$,$\rho_{6-9}\,$|$\,\varphi_2$>(Region<Pair<A,A>><$\toprgn$> r, 
                Pair<A,A><$\rho_{6-8}$> q, A<$\rho_{9}$> y) {
  q.fst := y; q.snd := y; 
  open r as $\rgn_0$ withroot p in
    letregion $\rgn_1$ in
      let x = new Pair<A,A><$\rgn_1$,$\rgn_0$,$\rgn_0$>
                      (p.fst,p.fst) in
        alt<$\rgn_{1}$,$\rgn_{1}$,$\rgn_{0}$,$\rgn_{0}$,$\rgn_{0}$>(r,x,p.snd)
}
\end{codejava}
% Note that, to avoid clutter, we have already resolved appropriate
% region arguments to the recursive call, instead of introducing new
% region variables and generating equality constraints on them. 
Constraints generated during the elaboration are shown below
(let $\rhoenv_2 = \{\rhoalloc_0,\rho_{0-4},\rhoalloc_2,\rho_{6-9}\}$ ):
\begin{smathpar}
\begin{array}{l}
\csid{7} \tywf{\rhoenv_2}{\varphi_2}\qquad
\csid{8} \isvalid{\varphi_1 \conj \varphi_2}
    {\rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6} \\
\csid{9} \isvalid{\varphi_1 \conj \varphi_2}{\rho_7 = \rho_9} \qquad
\csid{10} \isvalid{\varphi_1 \conj \varphi_2}{\rho_{8} = \rho_9} \\
\csid{11} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives
\rgn_1} 
    {[\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2}
\end{array}
\end{smathpar}
% The first constraint is generated by $\typeOk$ on the type of
% $\C{p}$.  Second and third are generated by the the assignment
% expressions, and the last constraint is generated from the recursive
% call.

@@

Our constraint generation algorithm traverses the entire program,
performing elaboration and collecting constraints. However, we do not
require region type inference to be a whole-program analysis.
Compositionality of type inference is reclaimed by solving the
constraints in a compositional fashion. Concretely, this means that
our constraint solving algorithm visits and solves every constraint
(or, every set of mutually dependent constraints) only once. It
composes computed solutions to solve other constraints that depend on
the solved constraints. Importantly, the failure to solve a dependent
constraint does not result in backtracking. We now describe our
constraint solving algorithm by demonstrating its peformance over the
example constraints shown in the previous section. A formal account of
the algorithm can be found in our technical report~\cite{techrep}.
% To simplify the presentation of our algorithm, we assume that are no
% mutually recursive definitions in the source program. Recursive
% definitions are nonetheless allowed.

\paragraph{Terminology} In a validity constraint, a predicate variable
occuring on the left side (resp. right side) of the turnstile is said
to occur negatively (resp. positively). A validity constraint
\emph{constrains} the set of predicate variables that occur negatively
in the constraint, while it \emph{uses} the set of predicate variables
that occur positively. A constraint is said to be \emph{recursive} if
it constrains and uses a predicate variable.
% A pair of validity constraints are said to be \emph{mutually
% dependent} if there exists a pair ($\varphi_1$, $\varphi_2$) of
% predicate variables that occur with opposite polarities in both the
% constraints. In such case, the pair ($\varphi_1$,$\varphi_2$) of
% predicate variables are also said to be mutually dependent. We call
% the transitive closure of mutual dependency relation as
% \emph{transitive dependency} relation. Like mutual dependency,
% transitive dependency is also extended to predicate variables.

Given a set of validity constraints, we first build a dependency graph
($G_c$) with constraints as nodes, and dependencies between them
captured as edges. There exists an edge from a constraint $c_2$ to a
constraint $c_1$ in the graph (i.e., $c_2$ \emph{depends on} $c_1$) if
any of the following conditions hold:
\begin{itemize}
\item \textbf{Type 1} $c_1$ constrains a predicate variable that $c_2$ uses. 
\item \textbf{Type 2} $c_1$ constrains a (non-strict) subset of predicate variables
that $c_2$ constrains.
\end{itemize}

% The first condition intuitively corresponds to a case, where
% expression or type, whose region elaboration is constrained by $c_2$
% refers to a method or a class, whose unknown precondition is
% constrained by $c_1$. The common predicate variable ($\varphi$)
% represents the unknown precondition in this case. The dependency from
% $c_2$ to $c_1$ means that $c_1$ must be solved to compute $\varphi$
% before $c_2$ is solved, thus enforcing the rule that the precondition
% of a method must not depend on its calling context. 
% Note that the first condition results in self-loops over recursive
% constraints in the dependency graph.  Since mutually recursive
% definitions refer each other, the first condition also results in
% bidirectional dependencies between mutually recursive constraints
% (and self-loops over recursive constraints). 
% The second condition captures two kinds of dependencies. First, the
% dependency from the constraints over a method precondition to the
% constraints over the precondition of the class containing the method.
% This captures our preference that the constraints over a class's
% region parameters should not depend on the idiosyncracies of its
% methods. Any additional constraints required by any of its methods
% must be captured in the precondition of the method itself
% (well-formedness rules allow this possibility). The second condition
% adds bidirectional dependencies between validity constraints that
% constraint the same set of predicate variables. 
Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.

\paragraph{Example} The dependency DAG ($G_C$) over validity
constraints from the \C{Pair} example (\S~\ref{sec:constraints}) is
shown in Fig.~\ref{fig:pair-deps}. Each node (labeled $C_i$) is a set
of constraints that belong to a strongly connected component in the
dependency graph ($G_c$), hence are mutually dependent. All
dependencies, except the self-dependency on $\mathbf{c_{10}}$, are
type-2 dependencies.

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. For the DAG in
Fig.~\ref{fig:pair-deps}, we consider the topological order $[C_1,
C_2, C_3]$, and solve the sets of constraints in that order.

% Consequently, when the turn of a constraint set ($C$) arrives during
% the constraint solving process, it satisfies certain properties:
% \begin{itemize}
% \item There exists only one predicate variable ($\varphi$) that is either
% constrained or used by the constraints in the set ($C$). The variable is
% called set's \emph{subject}. This property follows from (a). the fact
% that all the dependency constraints have already been solved (and
% solutions applied), and (b). the assumption that there are no mutually
% recursive definitions. 
% \item All the constraints that constrain the set's subject are present
% in the set. This follows from our definition of the dependency relation.
% \end{itemize}

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) varibles in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{\cite{techrep} formally justifies the soundness of this
step}. For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}
The constraint $c_{12}$ is a non-recursive constraint of the form
$\isvalid{\phictxt \conj \varphi}{\phicstr}$, where $\phictxt$ and
$\phicstr$ are concrete constraint formulas ($\phictxt$ is $true$ in
this case). Such constraints are solved for $\varphi$ by computing a
maximally weak $\phisol$ such that $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$. If such $\phisol$ satisfies the
well-formedness condition on $\varphi$, then it is clearly a solution
for $\varphi$, because there cannot exist a weaker solution for
$\varphi$ that implies $\phicstr$ under $\phictxt$. On contrary, if a
maximally weak $\phisol$ fails the well-formedness condition, then we
establish~\cite{techrep} that it is impossible to satisfy the
constraint. Hence, it suffices to obtain a maximally weak $\phisol$,
that satisfies $\isvalid{\phictxt}{\phisol \Leftrightarrow \phicstr}$
and check if it satisfies the well-formedness condition on $\varphi$.
The former problem can be cast as an instance of graph augmentation
problem, as described below.
\vspace*{-0.05in}
\paragraph{Definitions} Given a constraint formula $\phi$, we define
its graph encoding $G(\phi)=(V(\phi),E(\phi))$ as a digraph whose
vertices ($V(\phi)$) are free region variables and identifiers in
$\phi$, and whose edges ($E(\phi)$) denote outlives constraints
in\footnote{We alternate between viewing a constraint formula ($\phi$)
as a set of constraints and a conjunction of constraints in this
description} $\phi$.  That is, if $\phi$ contains a constraint $\rho_1
\outlives \rho_2$, then $\rho_1,\rho_2 \in V(\phi)$ and
$(\rho_1,\rho_2) \in E(\phi)$. Equality constraints are treated as a
conjunction of symmetric outlives constraints for the purpose of graph
encoding. Conversely, given a digraph $G$, we define its constraint
encoding $\Phi(G)$ in a straightforward manner. We say that a graph
$G_1=(V_1,E_1)$ is as connected as graph $G_2=(V_2,E_2)$ if $V_2
\subseteq V_1$, and for every $\rho_1, \rho_2 \in V_2$, if there
exists a path between $\rho_1$ and $\rho_2$ in $G_2$, then there must
exist a path between same pair of vertices in $G_1$.

A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
connected as $G(\phicstr)$. The problem of computing such a $G$ is
equivalent to the problem finding of finding minimum number of edges
to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
Efficient algorithms to solve the later problem are known to
exist~\cite{graph}, and we rely on them for the solution.
